Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-14 paper.tex
updated appendix

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/lum.stars.eps}
 \caption{\it Accumulative luminosity as a function of radius. We show
   the luminosity without substructures inside radius $r$. The solid
   lines show the optical/IR emission from stars and dust (SD) and the
   IC upscattered SD light is shown by the dashed lines. The red lines
   show the total emission including the brightest cluster galaxy, the
   intra cluster light and the additional intra cluster galaxies,
   while these galaxies have been cut out in the smooth component
   shown by the orange lines. We normalize the SD light and the IC
   upscattered SD with the smooth luminosity within $\rvir$ for each
   component. We find that the gamma-ray IC luminosities are dominated
   by the central regime, while the SD light is mainly build up in the
   outer parts of the cluster. For comparison, we show that the SD
   light traces the NFW mass profile (blue) outside $0.1\rvir$.}
 \label{fig:SD_lum}
\end{figure}

@@

In this work, we thoroughly study the spectral emission
characteristics of the DM emission. In general there are different
radiation mechanism that emit in the gamma-ray regime, namely (i)
continuum emission following the hadronization of the annihilating
neutralinos in benchmark models which leads to the production of
charged and neutral mesons which decay finally into electrons,
neutrinos, and gammas, (ii) inverse Compton (IC) emission by the
leptons in the final state which up-scatter radiation fields from the
cosmic microwave background, dust, and starlight, and (iii) final
state radiation or internal bremsstrahlung which dominates the highest
energy emission close to the rest mass of the self-annihilating
neutralinos (if present in the model under consideration). While this
final state radiation is sometimes not distinguished from the
continuum emission contribution, we keep it separate as it dominates
the very high-energy tail of the gamma-ray spectrum in LP models while
there the continuum  emission is negligibly low.

For the first time we compute the IC emission component not only from
the upscattered cosmic microwave (CMB) photons but also from a
realistically modeled photon field due to dust and starlight extending
from infra red to optical wavelengths. To this end, we construct a
rather simple analytic model for the spatial and spectral distribution
of dust and starlight in clusters which can be used for the
annihilation emission in galaxies with a different spatial profile.
We find that in benchmark models, the continuum emission dominates
over the IC emission of any available radiation field. In the LP
models, where the continuum emission is low, the IC up-scattering of
the CMB dominates below 60 GeV. For large clusters, the final state
radiation takes then over for energies up to the cutoff at $m_\chi c^2
\sim 1$~TeV. In small clusters, there is a window around $E_\gamma
\sim 100~\mathrm{GeV}$ where the IC up-scattering of the dust emission
dominates over the CMB IC emission and the final state radiation. The
IC emission of starlight remains always subdominant.  We stress, that
the inclusion of the dust and starlight radiation fields are
negligible in the cooling function for the leptons.  The main reason
why the IC emission from starlight and dust remains subdominant is the
small overlap of this component with the peripherally peaked
substructure mass density profiles. If for some reason, DM has less
substructure, the relative importance of this IC component increases
considerably.

@@

\section{Detailed modelling of starlight and dust in clusters}
\label{sect:SD}
In this section we show the details of our model of stars and dust
(SD) in clusters of galaxies. In the first part we use the spectral
distribution of the SD light in galaxies to derive a simple spectral
model for SD in galaxy clusters. This is done by renormalizing the
galactic SD light to match the conditions for a cluster. In the second
part we derive the spatial distirbution of the SD light.

\subsection{Energy spectrum}
The SD light from clusters is dominated by the emission from
galaxies. This radiation is expected to leak out into the much larger
volume of the ICM, hence we approximate the spectral distribution of
the ICL with the light from galaxies. We use the far infrared to
ultraviolet spectrum derived in \cite{2006ApJ...648L..29P} for a MW
type galaxy to characterize the SD spectral distribution of a
cluster. Note that we only keep the spectral shape, and renormalize
the spectral distribution using the luminosity from SD.

In Fig.~\ref{fig:SD_spectra} we show the spectral distribution of the
light from SD together with the CMB black-body distribution in a
$6\times10^{14}\,\msun$ galaxy cluster. We fit the spectral shape of
SD individually; the dust that peaks at about $10^{-2}\,\ev$ is fitted
using a double power-law, while the broader spectral distribution of
the stars that peaks at about $1\,\ev$ is fitted using a tipple
power-law. The best fitted SD spectra are given by: \del{both at the
  radius $r=0.03\rvir$ where the energy density of the smooth light
  from SD (see Fig.~\ref{fig:SD_Edens}) equals the energy density of
  the CMB. We have renormalized both the spectral data and the fitted
  spectra using the luminosity from stars and dust individually.}
\begin{eqnarray}
  u_\stars^\gal(\eph) &=& \frac{23\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{1.23\,\rmn{eV}}{\eph}\right)^{1.9} \nonumber \\
  &\times&\left[1+\left(\frac{2.04\,\rmn{eV}}{\eph}\right)^{20}\right]
  ^{-\frac{1.9}{20}}\nonumber \\
  &\times& \left[1+\left(\frac{0.78\,\rmn{eV}}{\eph}\right)^{20}\right]^{-\frac{2.6}{20}}\,, \\
  u_\dust^\gal(\eph) &=& 
  \frac{40\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{4.9}\nonumber \\
  &\times& \left[1+\left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{1.9}\right]^{-4.9}\,.
\end{eqnarray}
The specific energy density of SD light in a cluster can now be
derived from 
\begin{eqnarray} 
u_\sd(\eph, r) &=& j(r)\,u_\sd(\eph)\,,
\label{eq:u_SD_er}
\end{eqnarray}
where the SD spectrum is given by 
\begin{eqnarray}
  u_\sd(\eph) &\equiv& \eph^2\frac{\dd^2 N_\ph}{\dd \eph \dd V}
  =  \sum_i N_i(\mvir) u_i^\gal(\eph)\,,\nonumber \\ 
\rmn{where}\quad i&=&\{\rmn{stars,dust}\}\,.
\end{eqnarray}
Here $\eph$ denotes the energy of the SD photons, $j(r)$ describes the
unitless spatial distribution which we derive in
sect.~\ref{sect:SD_radial}, and $N_i(\mvir)$ denotes the mass
dependent normalization for the SD. This function is determined using
the total energy of the light in SD within $\rvir$ denoted by
$E_{i,\vir}$. We relate this energy to the luminosity $L_i$ of each
respective wavelength band (optical and IR light for the dust, and UV
light for the stars) using
\begin{eqnarray} 
  E_{i,\vir} &=& L_i \frac{\rvir}{c} \nonumber \\
  &=&N_i(\mvir)\int_{\rvir} \int_i \,j(r) 
  \frac{u_i^\gal(\eph)}{\eph}\,\dd V\dd \eph\,,\nonumber \\
 \rmn{for}\quad i&=&\{\rmn{stars,dust}\}\,.
\label{eq:E_SD}
\end{eqnarray}
We approximate the total energy of the photons within $\rvir$ with the
luminosity \cite{2008A&A...490..547G} times the typical timescale it
takes for a photon to propagate through the cluster (we assume that
the cluster is optically thin).

We derive the luminosity from the stars using
\begin{eqnarray}
L_\stars(M_{200\rmn{av}})&=&10^{\frac{m_\odot-m_\rmn{cl}-2.5\log\left(1+z\right)}{2.5}}
\left(\frac{D_\clu}{D_\odot}\right)^2 L_\odot\nonumber\\
&\approx& 5.3\times10^{44}\,\rmn{erg}\,\rmn{s}^{-1}\,,
\label{eq:L_stars}
\end{eqnarray}
where the average apparent magnitude of a cluster at $z\approx 0.25$
in the $(r+i)$-band is given by $m_\rmn{cl}\approx 15.5$
\cite{2005MNRAS.358..949Z}. This magnitude is derived from stacked
clusters observed by the SDSS with the average mass
$M_{200\rmn{av}}=4.0\times10^{13}\msun$ and redshift close to
$z=0.25$. We also use an apparent magnitude of the sun $m_\odot\approx
-27.7$, distance to the sun $D_\odot=4.85\times10^{-6}\,\rmn{pc}$,
luminosity of the sun $L_\odot=1.17\times10^{33}\,\rmn{erg\,s}^{-1}$,
and distance to the cluster $D_\clu=1.26\times10^9\,\rmn{pc}$. To
account for different cluster masses, we attach a simple mass scaling
to the luminosity from the stars in clusters in Eq.~(\ref{eq:L_stars}).
Since a large contribution of the total starlight in a cluster comes
from the brightest cluster galaxy (BCG), we assume that the total
starlight has the same halo mass scaling as the BCG
\cite{2010ApJ...713.1037H}
\begin{equation}
L_\stars(\mvir)=L_\stars(M_{200\rmn{av}})\,
\left(\frac{\mvir}{M_{200\rmn{av}}}\right)^{0.18}\,.
\label{eq:L_stars_m}
\end{equation} 
We can now fix the unitless normalization constant for the stars in
Eq.~(\ref{eq:E_SD}) by integrating over the cluster volume and
spectral distribution of the stars. It is given by
\begin{equation}
 N_\stars(\mvir) = 6.0\times10^{-9}\,\kpc^2\,
\left(\frac{\mvir}{M_{200\rmn{av}}}\right)^{0.18}\,
\frac{\rvir}{\int_{\rvir} j(r) \,\dd V}\,.
\label{eq:N_stars}
\end{equation}
\del{Note that since the luminosities include the contribution from
galaxies, we have integrated the spatial distribution including
galaxies to derive the normalization. Once the normalization is fixed,
the SD energy densities are derived from the spatial distribution
where galaxies have been excluded.}

The luminosity from the dust is derived from the luminosity-richness
scaling relation found in \cite{2008A&A...490..547G}
$L_\dust=10^{44.8}\left(\frac{N_{200}}{10}\right)^{0.8}$. We pick a
high richness cluster with $N_{200}=82$ to normalize the luminosity
since these clusters are less likely biased by chance
coincidences. This richness roughly correspond to a mass
$M_\rmn{200dust}=6.0\times10^{14}\,\msun$
\cite{2010ApJ...713.1037H}. Assuming that the dust scales with halo
mass in the same way as the stars, we get
\begin{equation}
L_\dust(\mvir)=L_\dust(M_\rmn{200dust})\,
\left(\frac{\mvir}{M_\rmn{200dust}}\right)^{0.18}\,.
\label{eq:L_dust_scale}
\end{equation} 
The normalization constant for the dust in Eq.~(\ref{eq:E_SD}) can now
be determined:
\begin{equation}
 N_\stars(\mvir) = 4.0\times10^{-7}\,\kpc^2\,
\left(\frac{\mvir}{M_\rmn{200dust}}\right)^{0.18}\,
\frac{\rvir}{\int_{\rvir} j(r) \,\dd V}\,.
\label{eq:N_dust}
\end{equation}

\subsection{Radial profiles}
\label{sect:SD_radial}
Our goal in this section is to derive a simple spatial model for the
distribution of SD in galaxy clusters. For this reason we use stacked
cluster observations by Sloan in the $r$ and $i$-band that trace the
stars and for simplicity we assume that the dust trace the stars in
the clusters \del{\bf WHY IS THIS A GOOD APPROXIMATION}. The stacked
surface brightness profiles in \cite{2005MNRAS.358..949Z} are measured
at redshift $z \sim 0.25$ with an average mass of the clusters of
$4.0\times10^{13}\,\msun$. The profiles are build up out of three
components; diffuse intracluster light (ICL), galaxies, and the
brightest cluster galaxy (BCG) in the center of clusters. However, the
total volume overlap of the ISM to the cluster is very small compared
to the ICL- and BCG overlap with the cluster, thus the relative
contribution of the galaxies to the IC emission is suppressed
(c.f. Eq.~\ref{eq:SD_overlap}). To remove this bias we cut out the
galaxies that have been smoothed over the cluster in the stacked
analysis. In Fig.~\ref{fig:SD_spatial} we show the SDSS stacked
brightness profiles as well as the fitted profiles\footnote{The
  measured brightness is converted into units of
  $\rmn{erg}\,\rmn{s}^{-1}$ using \cite{2010...book}
  $S(\rmn{mag}\,''^{-2}) =
  \mathcal{M}_\odot+21.6-2.5\log_{10}[S(L_\odot\,\rmn{pc}^{-2})]$,
  where the sun's absolute magnitude in the $r$-band is given by
  $\mathcal{M}_\odot=27.1$ \cite{1998gaas.book.....B} and the
  luminosity of the sun by $L_\odot=3.85\times10^{33}\,
  \rmn{erg}\,\rmn{s}^{-1}$.}, where both the data and fits are
normalized with the central surface brightness of the total SD
component ($S_\sd^\rmn{tot}(0)$). Our benchmark spatial model is shown
with the solid black line, and include the contribution from the ICL
and BCG.

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.photon.eps}
 \caption{\it Surface brightness profiles of the $r$-band ($h\nu\sim
   1\,\ev$) obtained from stacked clusters in the Sloan Digital Sky
   Survey (SDSS) at the redshift $z \sim 0.25$ \protect
   \cite{2005MNRAS.358..949Z}. The crosses show the total observed
   starlight including the diffuse intracluster (ICL), the galaxies,
   and the brightest cluster galaxy (BCG) in the center of the
   cluster. The diamonds denote the smooth part of the observed
   starlight from the ICL and the BCG. The solid line show the fit to
   the data of the total light including the ICL, the BCG, and the
   galaxies, while the smooth component is represented by the dashed
   line that is fitted to only the ICL and the BCG. Note that, we use
   total SD brightness in the cluster center ($S_\sd^\rmn{tot}(0)$) to
   normalize the SD surface brightness profiles.}
 \label{fig:SD_spatial}
\end{figure}

Instead of modelling the surface brightness with a de Vaucouleur
profile with the addition of a power-law, we use a double beta profile
model for simplicity of deprojection. It is given by
\begin{equation}
\tilde{S}_\sd (r_\bot) = \frac{S_\sd (r_\bot)}{S_\rmn{SD}^\rmn{tot}(0)} = 
\sum_{i=1}^2 \,\tilde{S}_i\, 
\left[1 + \left( \frac{r_\bot}{r_{\mathrm{c}_i}}\right)^2\right]
^{-3\beta_i + 1/2}\,,
\label{double_beta}
\end{equation}
where we found the following best fitted normalized central brightness
($S_i$), core radius ($r_{\mathrm{c}_i}$), and slope ($\beta_i$) for
the total model:
\begin{align}
&\tilde{S}_1^\rmn{tot} = 1.0\,,
&&r_{\mathrm{c}_1}^\rmn{tot} = 1.8\,\rmn{kpc}\,,
&&&\beta_1^\rmn{tot} = 0.45\, \nonumber\\
&\tilde{S}_2^\rmn{tot} = 2.3\times10^{-3}\,,
&&r_{\mathrm{c}_2}^\rmn{tot} = 0.19\,\rvir\,,
&&&\beta_2^\rmn{tot} = 0.44\,.\nonumber\\
& && &&&
\label{fit_spatial_IR}
\end{align}
and for the smooth model
\begin{align}
&\tilde{S}_1^\rmn{sm}=0.47,\,
&&r_{\mathrm{c}_1}^\rmn{sm}=3.9\,\rmn{kpc}\,,
&&&\beta_1^\rmn{sm}=0.53 \,\nonumber\\
&\tilde{S}_2^\rmn{sm}=8.3\times10^{-4}\,,
&&r_{\mathrm{c}_2}^\rmn{sm}=0.19\,\rvir\,,
&&&\beta_2^\rmn{sm}=0.54\,.\nonumber\\
& && &&&
\label{fit_spatial_IR_sm}
\end{align}
The three dimensional spatial profile is derived by deprojecting the
surface brightness in Eq.~(\ref{double_beta}) (see
e.g. \cite{2004A&A...413...17P} for details about the deprojection):
\begin{eqnarray}
  j(r)  = \sum_{i=1}^2 \frac{\tilde{S}_i}{2\pi\,r_{\mathrm{c}_i}}\,
  \frac{6 \beta_i - 1}{\left(1 + r^2/r^2_{\mathrm{c}_i}\right)^{3\beta_i}}\,
  \mathcal{B}\left(\frac{1}{2},3\beta_i\right)\,,\nonumber\\
\end{eqnarray}
where $\mathcal{B}(a,b)$ denotes the beta-function
\cite{1965hmfw.book.....A}.

The spatial distribution of the SD energy density is an important
quantity since it impacts the steady state electron and positron
distribution through cooling. A higher energy density of SD photons in
a cluster suppresses the gamma-ray flux from IC upscattered
photons. We derive the energy density from starlight and dust in a
galaxy cluster through
\begin{eqnarray}
\label{eq:u_SD_r}
u_\sd(r) &=& \int \dd \eph \frac{\dd^2 N_\ph}{\dd \eph \dd V}\,\eph
=\int \dd \eph \frac{u_\sd(\eph, r)}{\eph}
\nonumber \\
&=&  j(r)  \int \dd \eph \sum_i 
N_i(\mvir) \frac{u_i^\gal(\eph)}{\eph}\,, \nonumber \\
\end{eqnarray}
where we have used that specific energy density of the SD light given
by Eq.~(\ref{eq:u_SD_er}). Note that we compare the energy density
$u_\sd(r)$ to other radiation background fields in
Fig.~\ref{fig:SD_Edens} for a galaxy cluster with the mass
$\mvir=6\times10^{14}\msun$.

\del{ we compare the energy densities from different radiation fields
  in a galaxy cluster with the mass $\mvir=6\times10^{14}\msun$. We
  use two different profiles for the SD energy density, where the
  total profile includes the contribution from the ICL, the BCG and
  all the galaxies, while the galaxies are excluded in the smooth
  profile. We find that even for a cluster with a relative small mass
  the energy density of the SD components dominates over the CMB and
  the magnetic fields (with a central B field of 3~$\mu G$) within
  about 10\% of $\rvir$. Outside this radius the CMB is dominating the
  energy density of the cluster, hence the magnetic fields are always
  subdominant in the cluster and can safely be removed from cluster IC
  calculations for a reasonable central B-field and spectral
  index. Note that in this work we keep the contribution from B-fields
  to the total energy density for consistency.}

>>>>

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/lum.stars.eps}
 \caption{\it Accumulative luminosity as a function of radius. We show
   the luminosity without substructures inside radius $r$. The solid
   lines show the optical/IR emission from stars and dust (SD) and the
   IC upscattered SD light is shown by the dashed lines. The red lines
   show the total emission including the brightest cluster galaxy, the
   intra-cluster light and the additional intra-cluster galaxies,
   while these galaxies have been cut out in the smooth component
   shown by the orange lines. We normalize the SD light and the IC
   upscattered SD with the smooth luminosity within $\rvir$ for each
   component. We find that the gamma-ray IC luminosities are dominated
   by the central regime, while the SD light is mainly build up in the
   outer parts of the cluster. For comparison, we show that the SD
   light traces the NFW mass profile (blue) outside $0.1\rvir$.}
 \label{fig:SD_lum}
\end{figure}

@@

In this work, we thoroughly study the spectral emission
characteristics of the DM emission. In general there are different
radiation mechanism that emit in the gamma-ray regime, namely (i)
continuum emission following the hadronization of the annihilating
neutralinos in benchmark models which leads to the production of
charged and neutral mesons which decay finally into electrons,
neutrinos, and gammas, (ii) inverse Compton (IC) emission by the
leptons in the final state which up-scatter radiation fields from the
cosmic microwave background (CMB), dust, and starlight, and (iii) final
state radiation or internal bremsstrahlung which dominates the highest
energy emission close to the rest mass of the self-annihilating
neutralinos (if present in the model under consideration). While this
final state radiation is sometimes not distinguished from the
continuum emission contribution, we keep it separate as it dominates
the very high-energy tail of the gamma-ray spectrum in LP models while
there the continuum  emission is negligibly low.

For the first time we compute the IC emission component not only from
the upscattered CMB photons but also from a realistically modeled
photon field due to dust and starlight extending from infra red to
optical wavelengths. To this end, we construct a rather simple
analytic model for the spatial and spectral distribution of dust and
starlight in clusters which can be used for the annihilation emission
in galaxies with a different spatial profile.  We find that in
benchmark models, the continuum emission dominates over the IC
emission of any available radiation field. In the LP models, where the
continuum emission is low, the IC up-scattering of the CMB dominates
below 60 GeV. For large clusters, the final state radiation takes then
over for energies up to the cutoff at $m_\chi c^2 \sim 1$~TeV. In
small clusters, there is a window around $E_\gamma \sim
100~\mathrm{GeV}$ where the IC up-scattering of the dust emission
dominates over the CMB IC emission and the final state radiation. The
IC emission of starlight remains always subdominant.  We stress, that
the inclusion of the dust and starlight radiation fields are
negligible in the cooling function for the leptons.  The main reason
why the IC emission from starlight and dust remains subdominant is the
small overlap of this component with the peripherally peaked
substructure mass density profiles. If for some reason, DM has less
substructure, the relative importance of this IC component increases
considerably.

@@

\section{Detailed modelling of starlight and dust in clusters}
\label{sect:SD}
In this section we present the details of our model of stars and dust
(SD) in clusters of galaxies. In the first part we the measured
spectral distribution of light from SD in a typical galaxy to derive a
simple spectral model for SD in galaxy clusters. This is done by
renormalizing the galactic SD light to match the observed luminosity
from a cluster in the far infrared (IR) and ultraviolet (UV) bands. In
the second part we derive the spatial distribution of the SD light
from a stacked sample of clusters.

\subsection{Energy spectrum}
The SD light from clusters is dominated by the emission from
galaxies. This radiation is expected to leak out into the much larger
volume of the intra-cluster medium, hence we approximate the spectral
distribution of the intra-cluster light (ICL) with the light from
galaxies. We use the far IR to UV spectrum derived in reference
\cite{2006ApJ...648L..29P} for a Milky Way (MW) type galaxy to
characterize the spectral distribution of SD in a cluster. Note that
we only keep the spectral shape, and renormalize the spectral
distribution using the luminosity from SD in clusters.

In Fig.~\ref{fig:SD_spectra} we show the spectral distribution of the
light from SD together with the cosmic microwave background (CMB)
black-body distribution in a $6\times10^{14}\,\msun$ galaxy
cluster. We fit the spectral shape of SD individually; the dust that
peaks at about $10^{-2}\,\ev$ is fitted using a double power-law,
while the broader spectral distribution of the stars that peaks at
about $1\,\ev$ is fitted using a tipple power-law. The best fit to the
galactic spectrum is given by:
\begin{eqnarray}
  u_\stars^\gal(\eph) &=& \frac{23\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{1.23\,\rmn{eV}}{\eph}\right)^{1.9} \nonumber \\
  &\times&\left[1+\left(\frac{2.04\,\rmn{eV}}{\eph}\right)^{20}\right]
  ^{-\frac{1.9}{20}}\nonumber \\
  &\times& \left[1+\left(\frac{0.78\,\rmn{eV}}{\eph}\right)^{20}\right]^{-\frac{2.6}{20}}\,, \\
  u_\dust^\gal(\eph) &=& 
  \frac{40\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{4.9}\nonumber \\
  &\times& \left[1+\left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{1.9}\right]^{-4.9}\,.
\end{eqnarray}
The specific energy density of the light from SD in a cluster can now
be derived from
\begin{eqnarray} 
u_\sd(\eph, r) &=& j(r)\,u_\sd(\eph)\,,
\label{eq:u_SD_er}
\end{eqnarray}
where the spectral distribution of SD is given by
\begin{eqnarray}
  u_\sd(\eph) &\equiv& \eph^2\frac{\dd^2 N_\ph}{\dd \eph \dd V}
  =  \sum_i N_i(\mvir) u_i^\gal(\eph)\,,\nonumber \\ 
\rmn{for}\quad i&=&\{\rmn{stars,dust}\}\,.
\end{eqnarray}
Here $\eph$ denotes the energy of the SD photons, $j(r)$ describes the
unitless spatial distribution of SD (derived in
sect.~\ref{sect:SD_radial}), and $N_i(\mvir)$ denotes the mass
dependent normalization for the SD. This function is determined using
the total energy of the light from SD within $\rvir$ which is
represented by $E_{i,\vir}$. We relate this energy to the luminosity
$L_i$ of each respective wavelength band (optical and IR light for the
dust, and UV light for the stars) using
\begin{eqnarray} 
  E_{i,\vir} &=& L_i \frac{\rvir}{c} \nonumber \\
  &=&N_i(\mvir)\int_{\rvir} \int_i \,j(r) 
  \frac{u_i^\gal(\eph)}{\eph}\,\dd V\dd \eph\,,\nonumber \\
 \rmn{for}\quad i&=&\{\rmn{stars,dust}\}\,.
\label{eq:E_SD}
\end{eqnarray}
Here we approximate the total energy of the photons within $\rvir$
with the luminosity \cite{2008A&A...490..547G} multiplied the typical
timescale it takes for a photon to propagate through the cluster (we
assume that the cluster is optically thin).

We derive the {\em luminosity from the stars} using
\begin{eqnarray}
L_\stars(M_{200\rmn{av}})&=&10^{\frac{m_\odot-m_\rmn{cl}-2.5\log\left(1+z\right)}{2.5}}
\left(\frac{D_\clu}{D_\odot}\right)^2 L_\odot\nonumber\\
&\approx& 5.3\times10^{44}\,\rmn{erg}\,\rmn{s}^{-1}\,,
\label{eq:L_stars}
\end{eqnarray}
where the average apparent magnitude of a cluster at $z\approx 0.25$
in the $(r+i)$-band is given by $m_\rmn{cl}\approx 15.5$
\cite{2005MNRAS.358..949Z}. This magnitude is derived from stacked
clusters observed by the SDSS with the average mass
$M_{200\rmn{av}}=4.0\times10^{13}\msun$. We also use an apparent
magnitude of the sun $m_\odot\approx -27.7$, distance to the sun
$D_\odot=4.85\times10^{-6}\,\rmn{pc}$, luminosity of the sun
$L_\odot=1.17\times10^{33}\,\rmn{erg\,s}^{-1}$, and distance to the
cluster $D_\clu=1.26\times10^9\,\rmn{pc}$. To account for different
cluster masses, we attach a simple mass scaling to the luminosity from
the stars in clusters in Eq.~(\ref{eq:L_stars}).  Since a large
contribution of the total starlight in a cluster comes from the
brightest cluster galaxy (BCG), we assume that the total starlight has
the same halo mass scaling as the BCG \cite{2010ApJ...713.1037H}
\begin{equation}
L_\stars(\mvir)=L_\stars(M_{200\rmn{av}})\,
\left(\frac{\mvir}{M_{200\rmn{av}}}\right)^{0.18}\,.
\label{eq:L_stars_m}
\end{equation} 
We can now fix the unitless normalization constant for the stars in
Eq.~(\ref{eq:E_SD}) by integrating over the cluster volume and
spectral distribution of the stars:
\begin{equation}
 N_\stars(\mvir) = 
\left(\frac{\mvir}{M_{200\rmn{av}}}\right)^{0.18}\,
\frac{6.0\,\rvir\times10^{-9}\,\kpc^2}{\int_{\rvir} j(r) \,\dd V}\,.
\label{eq:N_stars}
\end{equation}

The {\em luminosity from the dust} is derived from the
luminosity-richness scaling relation found in
\cite{2008A&A...490..547G}
$L_\dust=10^{44.8}\left(\frac{N_{200}}{10}\right)^{0.8}$. We pick a
high richness cluster with $N_{200}=82$ to normalize the luminosity
since these clusters are less likely biased by chance
coincidences. This richness roughly correspond to a mass
$M_\rmn{200dust}=6.0\times10^{14}\,\msun$
\cite{2010ApJ...713.1037H}. We assume that the dust scales with halo
mass in the same way as the stars, and determine the normalization
constant for the dust in Eq.~(\ref{eq:E_SD}) to
\begin{equation}
 N_\stars(\mvir) = 
\left(\frac{\mvir}{M_\rmn{200dust}}\right)^{0.18}\,
\frac{4.0\,\rvir\times10^{-7}\,\kpc^2}{\int_{\rvir} j(r) \,\dd V}\,.
\label{eq:N_dust}
\end{equation}

\subsection{Radial profiles}
\label{sect:SD_radial}
Our goal in this section is to derive a simple spatial model for the
distribution of the light from SD in galaxy clusters. For this reason
we use stacked cluster observations by Sloan in the $r$ and $i$-band
that trace the stars and for simplicity we assume that the dust trace
the stars in the clusters {\bf Christoph, why is this a valid
  approximation?}. The stacked surface brightness profiles in
\cite{2005MNRAS.358..949Z} are measured at redshift $z \approx 0.25$
with an average mass of the clusters of $4.0\times10^{13}\,\msun$. The
profiles are build up out of three components; diffuse ICL, galaxies,
and the BCG in the center of clusters. However, the total volume
overlap of the inter-stellar medium with a cluster is very small
compared to the total overlap of the ICL- and BCG with a cluster, thus
the relative contribution of the galaxies to the IC emission is
suppressed (c.f. Eq.~\ref{eq:SD_overlap}). We remove this bias in our
benchmark model for SD by cutting out the galaxies which roughly
corresponds to $70$ percent of the total SD light. In
Fig.~\ref{fig:SD_spatial} we show the SDSS stacked brightness profiles
as well as the fitted profiles\footnote{The measured brightness is
  converted into units of $\rmn{erg}\,\rmn{s}^{-1}$ using
  \cite{2010...book} $S(\rmn{mag}\,''^{-2}) =
  \mathcal{M}_\odot+21.6-2.5\log_{10}[S(L_\odot\,\rmn{pc}^{-2})]$,
  where the sun's absolute magnitude in the $r$-band is given by
  $\mathcal{M}_\odot=27.1$ \cite{1998gaas.book.....B} and the
  luminosity of the sun by $L_\odot=3.85\times10^{33}\,
  \rmn{erg}\,\rmn{s}^{-1}$.}, where both the data and fits are
normalized with the central surface brightness of the total SD
component ($S_\sd^\rmn{tot}(0)$). Our benchmark spatial model is shown
with the solid black line, and include the contribution from the ICL
and BCG.

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.photon.eps}
 \caption{\it Surface brightness profiles of the $r$-band $(h\nu\sim
   1\,\ev)$ obtained from stacked clusters in the Sloan Digital Sky
   Survey (SDSS) at the redshift $z \sim 0.25$ \protect
   \cite{2005MNRAS.358..949Z}. The crosses show the total observed
   starlight including the diffuse intracluster (ICL), the galaxies,
   and the brightest cluster galaxy (BCG) in the center of the
   cluster. The diamonds denote the smooth part of the observed
   starlight from the ICL and the BCG. The solid line show the fit to
   the data of the total light including the ICL, the BCG, and the
   galaxies, while the smooth component is represented by the dashed
   line that is fitted to only the ICL and the BCG. Note that, we use
   total SD brightness in the cluster center
   $(S_\sd^\rmn{tot}(0))$ to normalize the SD surface
   brightness profiles.}
 \label{fig:SD_spatial}
\end{figure}

Instead of modelling the surface brightness with a de Vaucouleur
profile with the addition of a power-law, we use a double beta profile
model for simplicity of deprojection. It is given by
\begin{equation}
\tilde{S}_\sd (r_\bot) = \frac{S_\sd (r_\bot)}{S_\rmn{SD}^\rmn{tot}(0)} = 
\sum_{i=1}^2 \,\tilde{S}_i\, 
\left[1 + \left( \frac{r_\bot}{r_{\mathrm{c}_i}}\right)^2\right]
^{-3\beta_i + 1/2}\,,
\label{double_beta}
\end{equation}
where we found the following best fitted normalized central brightness
($S_i$), core radius ($r_{\mathrm{c}_i}$), and slope ($\beta_i$) for
the total model:
\begin{align}
&\tilde{S}_1^\rmn{tot} = 1.0\,,
&&r_{\mathrm{c}_1}^\rmn{tot} = 1.8\,\rmn{kpc}\,,
&&&\beta_1^\rmn{tot} = 0.45\, \nonumber\\
&\tilde{S}_2^\rmn{tot} = 2.3\times10^{-3}\,,
&&r_{\mathrm{c}_2}^\rmn{tot} = 0.19\,\rvir\,,
&&&\beta_2^\rmn{tot} = 0.44\,.\nonumber\\
& && &&&
\label{fit_spatial_IR}
\end{align}
For the smooth model, which is also our benchmark SD model, we find:
\begin{align}
&\tilde{S}_1^\rmn{sm}=0.47,\,
&&r_{\mathrm{c}_1}^\rmn{sm}=3.9\,\rmn{kpc}\,,
&&&\beta_1^\rmn{sm}=0.53 \,\nonumber\\
&\tilde{S}_2^\rmn{sm}=8.3\times10^{-4}\,,
&&r_{\mathrm{c}_2}^\rmn{sm}=0.19\,\rvir\,,
&&&\beta_2^\rmn{sm}=0.54\,.\nonumber\\
& && &&&
\label{fit_spatial_IR_sm}
\end{align}
The three dimensional spatial profile is derived by deprojecting the
surface brightness in Eq.~(\ref{double_beta}) (see
e.g. \cite{2004A&A...413...17P} for details about the deprojection):
\begin{eqnarray}
  j(r)  = \sum_{i=1}^2 \frac{\tilde{S}_i}{2\pi\,r_{\mathrm{c}_i}}\,
  \frac{6 \beta_i - 1}{\left(1 + r^2/r^2_{\mathrm{c}_i}\right)^{3\beta_i}}\,
  \mathcal{B}\left(\frac{1}{2},3\beta_i\right)\,,\nonumber\\
\end{eqnarray}
where $\mathcal{B}(a,b)$ denotes the beta-function
\cite{1965hmfw.book.....A}.

The spatial distribution of the energy density of SD is an important
quantity since it impacts the steady state electron and positron
distribution through cooling; a high energy density of SD photons in
a cluster suppresses the gamma-ray flux from IC upscattered
photons. We derive the energy density from starlight and dust in a
galaxy cluster through
\begin{eqnarray}
\label{eq:u_SD_r}
u_\sd(r) &=& \int \dd \eph \frac{\dd^2 N_\ph}{\dd \eph \dd V}\,\eph
=\int \dd \eph \frac{u_\sd(\eph, r)}{\eph}
\nonumber \\
&=&  j(r)  \int \dd \eph \sum_i 
N_i(\mvir) \frac{u_i^\gal(\eph)}{\eph}\,, \nonumber \\
\end{eqnarray}
where we have used that specific energy density of the SD light given
by Eq.~(\ref{eq:u_SD_er}). Note that we compare the energy density
$u_\sd(r)$ to other radiation background fields in
Fig.~\ref{fig:SD_Edens} for a galaxy cluster with the mass
$\mvir=6\times10^{14}\msun$. We find that inside a few percent of of
$\rvir$ the total energy density is dominated by the light from SD,
while the CMB is dominating elsewhere. This implies that if the boost
from dark matter (DM) substructures is significant then the overlap of
light from SD with the electron and positron distribution that trace
the substructures is small. Hence the resulting flux from inverse
Compton (IC) upscattered SD photons is suppressed compared to the IC
upscattered CMB photons.
